---
title: "Nominators' Behavior"
author: "Jonas Gehrlein @ Web3 Foundation"
date: 20/04/2023
output: html_document
version: "1.0"
---


```{r setup, fig.width=4, fig.height=3, warning=FALSE, message=FALSE, echo=FALSE}
# Load the required R packages:
library(plyr)
library(dplyr)
library(rmarkdown)
library(ggplot2)
library(reshape2)
library(tidyverse)
library(scales)
library(dbplyr)
library(assert)
library(ineq)
library(ggraph)
library(splitstackshape)
library(RCurl)
library(polkadotutils)
library(data.table)
```


```{r, message=FALSE}
# User Input
# Can be "polkadot" or "kusama"
chain = "polkadot"

# The sessions to analyze.
last_session = 6351
first_session = last_session - 544


# if we want not every era but every `time_offset` era. Using "1" means we include all eras. Using "2" would mean every second era etc.
time_offset <- 1
```

```{r, echo=FALSE}
# Parameters for Kusama
#first_session <- 27911-200
#last_session <- 27911+200
```

```{r, echo=FALSE, message=FALSE}
#Setting up the denomination KSM and (new) DOT:
network_identifier = ifelse(chain == "polkadot", "DOT", "KSM")
number_session_per_era <- 6

if(chain=="polkadot"){
  normalization = 1/10000000000
  chain_name = "Polkadot"
} else {
  normalization = 1/1000000000000
  chain_name = "Kusama"
}

# Additional normalization
normalization_thousands = 1/1000
normalization_millions = 1/1000000
normalization_billions = 1/1000000000

difference = last_session - first_session 
#x = c(last_session:(last_session - difference))

sessions = c(first_session:(first_session + difference))
# extract every time_offset's value
sessions <- sessions[seq(1, length(sessions), time_offset)]
old <- Sys.time() # get start time
```

```{r, echo=FALSE}
initiate_data <- function(df_t0, df_t1) {
  merged_df <- merge(df_t0, df_t1, all = TRUE, by = "stash_address", suffixes = c("_t0", "_t1"))
  
  merged_df$targets_changed <- vector("list", nrow(merged_df))
  merged_df$bonded_amount_changed <- vector("list", nrow(merged_df))
  merged_df$session_count <- 0
  
  for (i in 1:nrow(merged_df)) {
    if (is.na(merged_df$session_t0[i]) && !is.na(merged_df$session_t1[i])) {
      merged_df$targets_changed[[i]] <- c(merged_df$session_t1[i])
      merged_df$bonded_amount_changed[[i]] <- c(merged_df$session_t1[i])
      merged_df$session_count[i] <- 1
    } else if (!is.na(merged_df$stash_address[i]) && is.na(merged_df$session_t1[i])) {
      merged_df$targets_changed[[i]] <- c(merged_df$session_t0[i], df_t1$session[1])
      merged_df$bonded_amount_changed[[i]] <- c(merged_df$session_t0[i], df_t1$session[1])
      merged_df$session_count[i] <- 1
    } else {
      if (!is.na(merged_df$targets_t0[i]) && merged_df$targets_t0[i] != merged_df$targets_t1[i]) {
        merged_df$targets_changed[[i]] <- c(merged_df$session_t0[i], merged_df$session_t1[i])
      } else {
        merged_df$targets_changed[[i]] <- c(merged_df$session_t0[i])
      }
      
      if (!is.na(merged_df$bonded_amount_t0[i]) && merged_df$bonded_amount_t0[i] != merged_df$bonded_amount_t1[i]) {
        merged_df$bonded_amount_changed[[i]] <- c(merged_df$session_t0[i], merged_df$session_t1[i])
      } else {
        merged_df$bonded_amount_changed[[i]] <- c(merged_df$session_t0[i])
      }
      
      merged_df$session_count[i] <- 2
    }
  }
  
  final_df <- data.frame(stash_address = merged_df$stash_address,
                         targets = ifelse(is.na(merged_df$session_t1), NA, ifelse(is.na(merged_df$targets_t1), merged_df$targets_t0, merged_df$targets_t1)),
                         bonded_amount = ifelse(is.na(merged_df$session_t1), NA, ifelse(is.na(merged_df$bonded_amount_t1), merged_df$bonded_amount_t0, merged_df$bonded_amount_t1)),
                         targets_changed = I(merged_df$targets_changed),
                         bonded_amount_changed = I(merged_df$bonded_amount_changed),
                         session_count = merged_df$session_count,
                         session = merged_df$session_t1)
  
  return(final_df)
}
```

```{r, echo = FALSE}
# This function takes the output table of `initiate_data` and updates it with the information of the next session data. The function checks whether the bonded_amount and targets entry for each nominator did change between the current session and the next session. If so, it adds an entry of which session those variables changed to the respective variables.
update_data <- function(initial_data, new_data) {
  merged_df <- merge(initial_data, new_data, all = TRUE, by = "stash_address", suffixes = c("_old", "_new"))
  
  for (i in seq_len(nrow(merged_df))) {
    if (!is.na(merged_df$session_old[i]) && !is.na(merged_df$session_new[i])) {
      if (merged_df$targets_old[i] != merged_df$targets_new[i]) {
        merged_df$targets_changed[[i]] <- append(merged_df$targets_changed[[i]], merged_df$session_new[i])
      }
      if (merged_df$bonded_amount_old[i] != merged_df$bonded_amount_new[i]) {
        merged_df$bonded_amount_changed[[i]] <- append(merged_df$bonded_amount_changed[[i]], merged_df$session_new[i])
      }
      merged_df$session_count[i] <- merged_df$session_count[i] + 1
    } else if (is.na(merged_df$session_old[i]) && !is.na(merged_df$session_new[i])) {
      merged_df$targets_changed[[i]] <- c(merged_df$session_new[i])
      merged_df$bonded_amount_changed[[i]] <- c(merged_df$session_new[i])
      merged_df$session_count[i] <- 1
    } else if (!is.na(merged_df$session_old[i]) && is.na(merged_df$session_new[i])) {
      merged_df$targets_changed[[i]] <- append(merged_df$targets_changed[[i]], new_data$session[1])
      merged_df$bonded_amount_changed[[i]] <- append(merged_df$bonded_amount_changed[[i]], new_data$session[1])
    }
  }
  
  final_df <- data.frame(stash_address = merged_df$stash_address,
                         targets = ifelse(is.na(merged_df$session_new), merged_df$targets_old, merged_df$targets_new),
                         bonded_amount = ifelse(is.na(merged_df$session_new), merged_df$bonded_amount_old, merged_df$bonded_amount_new),
                         targets_changed = I(merged_df$targets_changed),
                         bonded_amount_changed = I(merged_df$bonded_amount_changed),
                         session_count = merged_df$session_count,
                         session = merged_df$session_new,
                         stringsAsFactors = FALSE)
  
  return(final_df)
}
``` 

```{r, echo=FALSE}
missing_tables <- 0
nominators_found <- TRUE

# To get a vector of sessions that were found (excluding those that have no data)
sessions_exist <- sessions

assert(url.exists(paste("https://storage.googleapis.com/watcher-csv-exporter/", chain , "_nominators_session_", first_session, ".csv", sep=(""))))
# Getting the first data frame to start the process
nominators_t0 <- getStakingData("nominators", "session", sessions[1], chain, "online")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Starting the loop at index 2 since we already got the first data frame manually.
for(i in 2:length(sessions)) {
  message(paste0("loop in session ", sessions[i] ))

  tryCatch(
  nominators_next <- read.csv(paste("~/Documents/Projects/staking-data/", chain , "_nominators_session_", sessions[i], ".csv", sep=("")), encoding = "UTF-8"),
  error = function(e){
    nominators_found <<- FALSE
    message(paste0("error in ", sessions[i]))
    })
  
  # One additional check if the nominators_next table is good. For example, session 5779 exists but the table is empty, which causes every nominator to get a change variable wrongly. With this, we skip empty tables (that are very rare it seems).
  if(is.na(nominators_next$era[1])){
    nominators_found = FALSE
  }
   
  if(nominators_found == FALSE){
      # Remove era that is missing from the eras list
      sessions_exist <- sessions[! sessions %in% c(sessions[i])]
      message(paste0("Session ", sessions[i], " missing, skipping"))
      # Track missing tables
      missing_tables <- missing_tables + 1
      # Reset flags
      nominators_found = TRUE
      # Skip rest of the execution and go for next iteration 
      next
  }
  if(i == 2){
    overall_nominators <- initiate_data(nominators_t0, nominators_next)
    } else {
    overall_nominators <- update_data(overall_nominators, nominators_next)
    }
}

# print elapsed time
new <- Sys.time() - old # calculate difference
print(new) # print in nice format
```

```{r, echo=FALSE}
# ------------------------ General data manipulation --------------------------

## Bonded Amount changes
# Getting the number of entries in the list to know how many changes occurred
for(i in 1:nrow(overall_nominators)){
  overall_nominators$nr_amount_changed[i] <- length(overall_nominators$bonded_amount_changed[i][[1]])
}

# Getting the number of entries in the list to know how many changes occurred
for(i in 1:nrow(overall_nominators)){
  overall_nominators$nr_targets_changed[i] <- length(overall_nominators$targets_changed[i][[1]])
}

overall_nominators$total_changes <- overall_nominators$nr_amount_changed + overall_nominators$nr_targets_changed
```

```{r echo = FALSE}
## Analysis Targets Change

# Create a dataframe that unnests the list of sessions per nominator and aggregates the number of changes on session level
targets_changed_per_session <- overall_nominators %>%
  select(stash_address, targets_changed) %>%
  unnest(cols = targets_changed)
targets_changed_per_session$tracker <- 1
targets_changed_per_session_aggregated <- aggregate(targets_changed_per_session$tracker, by = list(targets_changed_per_session$targets_changed), FUN = sum)
colnames(targets_changed_per_session_aggregated) <- c("session", "number_changes")

# There are a few missing tables which means that changes "aggregate" over a longer period of time until a new data set is found and then all the changes are accounted to that new session. To smoothen this out, I take the average change per session if there are missing tables. 
for(i in 1:nrow(targets_changed_per_session_aggregated)){
  if(i == 1){
    targets_changed_per_session_aggregated$distance[i] <- 1
  } else {
    targets_changed_per_session_aggregated$distance[i] <- targets_changed_per_session_aggregated$session[i] - targets_changed_per_session_aggregated$session[i-1]
  }
}
targets_changed_per_session_aggregated$average_nr_changes <- targets_changed_per_session_aggregated$number_changes / targets_changed_per_session_aggregated$distance
# The percentage of changes in the targets of a nominator. Since we count the first appearance of a nominator as a change in target, we substract it here. And since this is session data, a reasonable time frame to change something would be once per era at maximum. 
overall_nominators$targets_changed_percent <- (overall_nominators$nr_targets_changed - 1) / (overall_nominators$session_count/number_session_per_era)
```

```{r, echo = FALSE}
# Analysis Bonded Amount Change
bonded_changed_per_session <- overall_nominators %>%
  select(stash_address, bonded_amount_changed) %>%
  unnest(cols = bonded_amount_changed)

bonded_changed_per_session$tracker <- 1
bonded_changed_per_session_aggregated <- aggregate(bonded_changed_per_session$tracker, by = list(bonded_changed_per_session$bonded_amount_changed), FUN = sum)
colnames(bonded_changed_per_session_aggregated) <- c("session", "number_changes")

for(i in 1:nrow(bonded_changed_per_session_aggregated)){
  if(i == 1){
    bonded_changed_per_session_aggregated$distance[i] <- 1
  } else {
    bonded_changed_per_session_aggregated$distance[i] <- bonded_changed_per_session_aggregated$session[i] - bonded_changed_per_session_aggregated$session[i-1]
  }
}
bonded_changed_per_session_aggregated$average_nr_changes <- bonded_changed_per_session_aggregated$number_changes / bonded_changed_per_session_aggregated$distance
# The percentage of changes in the bonds of a nominator. Since we count the first appearance of a nominator as a change in bonds, we subtract it here. And since this is session data, a reasonable time frame to change something would be once per era at maximum. 
overall_nominators$bonded_changed_percent <- (overall_nominators$nr_amount_changed - 1) / (overall_nominators$session_count/number_session_per_era)
```

# Introduction

This analysis gives insights into the behavior of nominators within the `r chain` network. Nominators are tracked across consecutive sessions using their unique stash address. Within this dataset, each session provides insight into the `bonded_amount` and a list of targets, representing the nominators' selected nominations. This approach enables a comparison between sessions, effectively monitoring any changes that occur over time.

The following table illustrates the parameters of the analysis. Note, that there have been some outages in the data collection, but the analysis copes with that and minimizes the impact on the outcome quality.
<table style="border-collapse:collapse;border-spacing:0" class="tg"><thead><tr><th style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;font-weight:normal;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">Parameter</th><th style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;font-weight:normal;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">Value</th></tr></thead><tbody><tr><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">Chain</td><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">`r chain`</td></tr><tr><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">First Session</td><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">`r first_session`</td></tr><tr><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">Last Session</td><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">`r last_session`</td></tr><tr><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">Missing Tables</td><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal"><span style="color:#905;background-color:#ddd">`r missing_tables`</span></td></tr><tr><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">Total number of unique nominators</td><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal"><span style="color:#905;background-color:#ddd">`r nrow(overall_nominators)`</span></td></tr></tbody></table>

# Changes over time
In this section, we examine the temporal fluctuations in both the bonded amount and the targets chosen by each nominator. This analysis helps to identify trends and patterns in the behavior of nominators over time.

### Bonded Amount: Histogram of changes

The analysis primarily emphasizes the frequency of alterations in the bonded amount. It is essential to note that changes only take effect following a new election, which occurs every `r number_session_per_era` sessions. To account for this, it is more meaningful to normalize the frequency on a "per era" basis rather than per session. In this context, if a nominator modifies their bonded amount once throughout `r number_session_per_era` sessions, the frequency in the histogram is represented as 1. Conversely, if a nominator adjusts their bonded_amount every session, the frequency is denoted as 6.

```{r, echo = FALSE}
# Create histogram with ggplot2
ggplot(overall_nominators, aes(x = bonded_changed_percent)) +
  geom_histogram(binwidth = 1, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Bonded Amount Changed Percentage",
       x = "Bonded Amount Changed Percentage", y = "Frequency")
```

There are a few outlier nominators who adjust their bonded amount with remarkable frequency. For the purpose of this analysis, we will concentrate on those with a frequency lying between 0 and 1, ensuring a more representative understanding of typical nominator behavior.

```{r, echo = FALSE}
# Create histogram with ggplot2
ggplot(subset(overall_nominators, bonded_changed_percent <= 1), aes(x = bonded_changed_percent)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Bonded Amount Changed Percentage",
       x = "Bonded Amount Changed Percentage", y = "Frequency")
```

### Bonded amount: Changes over sessions

The subsequent graph illustrates the variations in bonded amount across sessions. The frequency displayed is in absolute terms, representing the total number of nominators who altered their bonded_amount within each individual session. This visualization helps to convey the overall trends and patterns in nominators' behavior over time.

```{r, echo = FALSE}
ggplot(subset(bonded_changed_per_session_aggregated, session != first_session), aes(x = session, y = average_nr_changes)) +
  geom_line() +
  geom_point(size = 0.5) +
  theme_minimal() +
  labs(title = "Number of Changes Over Time (Bonded Amount Change)",
       x = "Session",
       y = "Number of Changes") +
  theme(panel.grid.minor = element_blank())
```

We can observe a high number of changes in the bonded amount. However, this is primarily due to many nominators employing automatic re-bonding of staking rewards. Distinguishing manual changes in the bonded amount from automatic ones is challenging. An approximation could be made based on the stash address size and the expected staking rewards per session, but this is left unexplored for now.

## Targets
This analysis forms the most crucial aspect of nominator behavior as it directly sheds light on their validator preferences and the frequency with which they update their nominations. The following histogram displays individual changes in targets across nominators over the entire time frame. The data is normalized by the number of sessions a nominator has been active and is presented on a per-era basis. In other words, a value of 1 signifies that a nominator altered their targets once every six sessions they were active. It is important to note that, in contrast to the `bonded_amount`, changes in targets are always manually initiated by nominators, reflecting conscious decision-making.

### Targets: Histogram of changes

```{r, echo = FALSE}
# Create histogram with ggplot2
ggplot(overall_nominators, aes(x = targets_changed_percent)) +
  geom_histogram(binwidth = 1, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Targets Changed Percentage",
       x = "Targets Changed Percentage", y = "Frequency")
```

Once more, we encounter some outlier nominators who alter their targets with exceptional frequency, which hampers the graph's readability. To address this, the following graph narrows its focus to nominators who modify their targets between 0 and 100% of the time, specifically every six sessions. This refined visualization aids in better understanding the typical behavior of nominators in relation to their target changes.

```{r, echo = FALSE}
# Create histogram with ggplot2
ggplot(subset(overall_nominators, targets_changed_percent <= 1), aes(x = targets_changed_percent)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Targets Changed Percentage",
       x = "Targets Changed Percentage", y = "Frequency")
```

The following graph plots the frequency of target changes of nominators per session.

```{r, echo = FALSE}
ggplot(subset(targets_changed_per_session_aggregated, session != first_session), aes(x = session, y = average_nr_changes)) +
  geom_line() +
  geom_point(size = 0.5) +
  theme_minimal() +
  labs(title = "Number of Changes Over Time (Targets Change)",
       x = "Session",
       y = "Number of Changes") +
  theme(panel.grid.minor = element_blank())
```



# Nomination of inactive validators

```{r, echo = FALSE}
newest_nominators <- polkadotutils::getStakingData("nominators", "session", last_session, chain, "online")
newest_validators <- polkadotutils::getStakingData("validators", "session", last_session, chain, "online")

# Unnest the targets column for each nominator and create a new data frame that has a row for each entry in "targets".
nominators_unnested <- subset(newest_nominators, select = c("stash_address", "targets")) %>%
    separate_rows(targets, sep = ",")

colnames(nominators_unnested) <- c("nominator_stash", "stash_address")

merged <- merge(nominators_unnested, subset(newest_validators, select = c("stash_address", "active")), by = "stash_address", all = TRUE)
merged$counter <- 1

merged_aggregated <- aggregate(cbind(active, counter) ~ nominator_stash, data = merged, FUN = sum, na.rm = TRUE)

merged_aggregated$ratio <- merged_aggregated$active / merged_aggregated$counter
merged_aggregated$at_least_one_inactive <- ifelse(merged_aggregated$ratio < 1, 1,0)
  
```

The subsequent histogram presents the ratio of active to inactive validators within the nominators' targets. Unlike previous analyses, this examination focuses on the current state of the network during the most recent session (`r last_session`). In this context, a ratio of 1 indicates that all validators a nominator is voting for are presently active. This insight offers an understanding of the overall alignment of nominator preferences with the active validator pool.

```{r, echo=FALSE}
ggplot(merged_aggregated, aes(x = ratio)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "lightblue") +
  labs(title = "Histogram of ratio of nominations of active validators",
       x = "Ratio of nominations of active validators", y = "Frequency")
```

In general, there are `r format(sum(merged_aggregated$at_least_one_inactive), scientific = FALSE)` (or `r round(sum(merged_aggregated$at_least_one_inactive) / nrow(merged_aggregated)*100, digits=2)`%) nominators that have at least one inactive validator in their targets.
