---
title: "Nominators' Behavior"
author: "Jonas Gehrlein @ Web3 Foundation"
date: 27/03/2023
output: html_document
version: "1.6"
---


```{r setup, fig.width=4, fig.height=3, warning=FALSE, message=FALSE, echo=FALSE}
# Load the required R packages:
library(plyr)
library(dplyr)
library(rmarkdown)
library(ggplot2)
library(reshape2)
library(tidyverse)
library(scales)
library(dbplyr)
library(assert)
library(ineq)
library(ggraph)
library(splitstackshape)
library(RCurl)
library(polkadotutils)
library(data.table)
```


```{r, message=FALSE}
# User Input
# Can be "polkadot" or "kusama"
chain = "polkadot"

# The eras to analyze.
first_session = 6304 - 542
#first_session <- 27911-200
#last_session <- 27911+200
last_session = 6304


# if we want not every era but every `time_offset` era. Using "1" means we include all eras. Using "2" would mean every second era etc.
time_offset <- 1
```

```{r, echo=FALSE, message=FALSE}
#Setting up the denomination KSM and (new) DOT:
network_identifier = ifelse(chain == "polkadot", "DOT", "KSM")
number_session_per_era <- 6

if(chain=="polkadot"){
  normalization = 1/10000000000
  chain_name = "Polkadot"
} else {
  normalization = 1/1000000000000
  chain_name = "Kusama"
}

# Additional normalization
normalization_thousands = 1/1000
normalization_millions = 1/1000000
normalization_billions = 1/1000000000

difference = last_session - first_session 
#x = c(last_session:(last_session - difference))

sessions = c(first_session:(first_session + difference))
# extract every time_offset's value
sessions <- sessions[seq(1, length(sessions), time_offset)]
old <- Sys.time() # get start time
```

```{r, echo=FALSE}
initiate_data <- function(df_t0, df_t1) {
  merged_df <- merge(df_t0, df_t1, all = TRUE, by = "stash_address", suffixes = c("_t0", "_t1"))
  
  merged_df$targets_changed <- vector("list", nrow(merged_df))
  merged_df$bonded_amount_changed <- vector("list", nrow(merged_df))
  merged_df$session_count <- 0
  
  for (i in 1:nrow(merged_df)) {
    if (is.na(merged_df$session_t0[i]) && !is.na(merged_df$session_t1[i])) {
      merged_df$targets_changed[[i]] <- c(merged_df$session_t1[i])
      merged_df$bonded_amount_changed[[i]] <- c(merged_df$session_t1[i])
      merged_df$session_count[i] <- 1
    } else if (!is.na(merged_df$stash_address[i]) && is.na(merged_df$session_t1[i])) {
      merged_df$targets_changed[[i]] <- c(merged_df$session_t0[i], df_t1$session[1])
      merged_df$bonded_amount_changed[[i]] <- c(merged_df$session_t0[i], df_t1$session[1])
      merged_df$session_count[i] <- 1
    } else {
      if (!is.na(merged_df$targets_t0[i]) && merged_df$targets_t0[i] != merged_df$targets_t1[i]) {
        merged_df$targets_changed[[i]] <- c(merged_df$session_t0[i], merged_df$session_t1[i])
      } else {
        merged_df$targets_changed[[i]] <- c(merged_df$session_t0[i])
      }
      
      if (!is.na(merged_df$bonded_amount_t0[i]) && merged_df$bonded_amount_t0[i] != merged_df$bonded_amount_t1[i]) {
        merged_df$bonded_amount_changed[[i]] <- c(merged_df$session_t0[i], merged_df$session_t1[i])
      } else {
        merged_df$bonded_amount_changed[[i]] <- c(merged_df$session_t0[i])
      }
      
      merged_df$session_count[i] <- 2
    }
  }
  
  final_df <- data.frame(stash_address = merged_df$stash_address,
                         targets = ifelse(is.na(merged_df$session_t1), NA, ifelse(is.na(merged_df$targets_t1), merged_df$targets_t0, merged_df$targets_t1)),
                         bonded_amount = ifelse(is.na(merged_df$session_t1), NA, ifelse(is.na(merged_df$bonded_amount_t1), merged_df$bonded_amount_t0, merged_df$bonded_amount_t1)),
                         targets_changed = I(merged_df$targets_changed),
                         bonded_amount_changed = I(merged_df$bonded_amount_changed),
                         session_count = merged_df$session_count,
                         session = merged_df$session_t1)
  
  return(final_df)
}
```

```{r, echo = FALSE}
# This function takes the output table of `initiate_data` and updates it with the information of the next session data. The function checks whether the bonded_amount and targets entry for each nominator did change between the current session and the next session. If so, it adds an entry of which session those variables changed to the respective variables.
update_data <- function(initial_data, new_data) {
  merged_df <- merge(initial_data, new_data, all = TRUE, by = "stash_address", suffixes = c("_old", "_new"))
  
  for (i in seq_len(nrow(merged_df))) {
    if (!is.na(merged_df$session_old[i]) && !is.na(merged_df$session_new[i])) {
      if (merged_df$targets_old[i] != merged_df$targets_new[i]) {
        merged_df$targets_changed[[i]] <- append(merged_df$targets_changed[[i]], merged_df$session_new[i])
      }
      if (merged_df$bonded_amount_old[i] != merged_df$bonded_amount_new[i]) {
        merged_df$bonded_amount_changed[[i]] <- append(merged_df$bonded_amount_changed[[i]], merged_df$session_new[i])
      }
      merged_df$session_count[i] <- merged_df$session_count[i] + 1
    } else if (is.na(merged_df$session_old[i]) && !is.na(merged_df$session_new[i])) {
      merged_df$targets_changed[[i]] <- c(merged_df$session_new[i])
      merged_df$bonded_amount_changed[[i]] <- c(merged_df$session_new[i])
      merged_df$session_count[i] <- 1
    } else if (!is.na(merged_df$session_old[i]) && is.na(merged_df$session_new[i])) {
      merged_df$targets_changed[[i]] <- append(merged_df$targets_changed[[i]], new_data$session[1])
      merged_df$bonded_amount_changed[[i]] <- append(merged_df$bonded_amount_changed[[i]], new_data$session[1])
    }
  }
  
  final_df <- data.frame(stash_address = merged_df$stash_address,
                         targets = ifelse(is.na(merged_df$session_new), merged_df$targets_old, merged_df$targets_new),
                         bonded_amount = ifelse(is.na(merged_df$session_new), merged_df$bonded_amount_old, merged_df$bonded_amount_new),
                         targets_changed = I(merged_df$targets_changed),
                         bonded_amount_changed = I(merged_df$bonded_amount_changed),
                         session_count = merged_df$session_count,
                         session = merged_df$session_new,
                         stringsAsFactors = FALSE)
  
  return(final_df)
}
``` 

```{r, echo=FALSE}
missing_tables <- 0
nominators_found <- TRUE

# To get a vector of sessions that were found (excluding those that have no data)
sessions_exist <- sessions

assert(url.exists(paste("https://storage.googleapis.com/watcher-csv-exporter/", chain , "_nominators_session_", first_session, ".csv", sep=(""))))
# Getting the first data frame to start the process
nominators_t0 <- getStakingData("nominators", "session", sessions[1], chain, "online")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Starting the loop at index 2 since we already got the first data frame manually.
for(i in 2:length(sessions)) {
  message(paste0("loop in session ", sessions[i] ))

  tryCatch(
  nominators_next <- read.csv(paste("~/Documents/Projects/staking-data/", chain , "_nominators_session_", sessions[i], ".csv", sep=("")), encoding = "UTF-8"),
  error = function(e){
    nominators_found <<- FALSE
    message(paste0("error in ", sessions[i]))
    })
  
  # One additional check if the nominators_next table is good. For example, session 5779 exists but the table is empty, which causes every nominator to get a change variable wrongly. With this, we skip empty tables (that are very rare it seems).
  if(is.na(nominators_next$era[1])){
    nominators_found = FALSE
  }
   
  if(nominators_found == FALSE){
      # Remove era that is missing from the eras list
      sessions_exist <- sessions[! sessions %in% c(sessions[i])]
      message(paste0("Session ", sessions[i], " missing, skipping"))
      # Track missing tables
      missing_tables <- missing_tables + 1
      # Reset flags
      nominators_found = TRUE
      # Skip rest of the execution and go for next iteration 
      next
  }
  if(i == 2){
    overall_nominators <- initiate_data(nominators_t0, nominators_next)
    } else {
    overall_nominators <- update_data(overall_nominators, nominators_next)
    }
}

# print elapsed time
new <- Sys.time() - old # calculate difference
print(new) # print in nice format
```

```{r, echo=FALSE}
# ------------------------ General data manipulation --------------------------

## Bonded Amount changes
# Getting the number of entries in the list to know how many changes occurred
for(i in 1:nrow(overall_nominators)){
  overall_nominators$nr_amount_changed[i] <- length(overall_nominators$bonded_amount_changed[i][[1]])
}

# Getting the number of entries in the list to know how many changes occurred
for(i in 1:nrow(overall_nominators)){
  overall_nominators$nr_targets_changed[i] <- length(overall_nominators$targets_changed[i][[1]])
}

overall_nominators$total_changes <- overall_nominators$nr_amount_changed + overall_nominators$nr_targets_changed
```

```{r echo = FALSE}
## Analysis Targets Change

# Create a dataframe that unnests the list of sessions per nominator and aggregates the number of changes on session level
targets_changed_per_session <- overall_nominators %>%
  select(stash_address, targets_changed) %>%
  unnest(cols = targets_changed)
targets_changed_per_session$tracker <- 1
targets_changed_per_session_aggregated <- aggregate(targets_changed_per_session$tracker, by = list(targets_changed_per_session$targets_changed), FUN = sum)
colnames(targets_changed_per_session_aggregated) <- c("session", "number_changes")

# There are a few missing tables which means that changes "aggregate" over a longer period of time until a new data set is found and then all the changes are accounted to that new session. To smoothen this out, I take the average change per session if there are missing tables. 
for(i in 1:nrow(targets_changed_per_session_aggregated)){
  if(i == 1){
    targets_changed_per_session_aggregated$distance[i] <- 1
  } else {
    targets_changed_per_session_aggregated$distance[i] <- targets_changed_per_session_aggregated$session[i] - targets_changed_per_session_aggregated$session[i-1]
  }
}
targets_changed_per_session_aggregated$average_nr_changes <- targets_changed_per_session_aggregated$number_changes / targets_changed_per_session_aggregated$distance
# The percentage of changes in the targets of a nominator. Since we count the first appearance of a nominator as a change in target, we substract it here. And since this is session data, a reasonable time frame to change something would be once per era at maximum. 
overall_nominators$targets_changed_percent <- (overall_nominators$nr_targets_changed - 1) / (overall_nominators$session_count/number_session_per_era)
```

```{r, echo = FALSE}
# Analysis Bonded Amount Change
bonded_changed_per_session <- overall_nominators %>%
  select(stash_address, bonded_amount_changed) %>%
  unnest(cols = bonded_amount_changed)

bonded_changed_per_session$tracker <- 1
bonded_changed_per_session_aggregated <- aggregate(bonded_changed_per_session$tracker, by = list(bonded_changed_per_session$bonded_amount_changed), FUN = sum)
colnames(bonded_changed_per_session_aggregated) <- c("session", "number_changes")

for(i in 1:nrow(bonded_changed_per_session_aggregated)){
  if(i == 1){
    bonded_changed_per_session_aggregated$distance[i] <- 1
  } else {
    bonded_changed_per_session_aggregated$distance[i] <- bonded_changed_per_session_aggregated$session[i] - bonded_changed_per_session_aggregated$session[i-1]
  }
}
bonded_changed_per_session_aggregated$average_nr_changes <- bonded_changed_per_session_aggregated$number_changes / bonded_changed_per_session_aggregated$distance
# The percentage of changes in the bonds of a nominator. Since we count the first appearance of a nominator as a change in bonds, we subtract it here. And since this is session data, a reasonable time frame to change something would be once per era at maximum. 
overall_nominators$bonded_changed_percent <- (overall_nominators$nr_amount_changed - 1) / (overall_nominators$session_count/number_session_per_era)
```

# Introduction

This analysis investigates the behavior of nominators in the `r chain` network. To do so, their behavior can be tracked over consecutive sessions where they are uniquely identified by their `stash address`. In this dataset and for each session, we can observer their `bonded_amount` as well as a list of `targets` (i.e., their nominations). This allows to compare each session to the next and track changes that occur.

The following table illustrates the parameters of the analysis. Note, that there have been some outages in the data collection, but the analysis copes with that and minimizes the impact on the outcome quality.
<table style="border-collapse:collapse;border-spacing:0" class="tg"><thead><tr><th style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;font-weight:normal;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">Parameter</th><th style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;font-weight:normal;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">Value</th></tr></thead><tbody><tr><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">Chain</td><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">`r chain`</td></tr><tr><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">First Session</td><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">`r first_session`</td></tr><tr><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">Last Session</td><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">`r last_session`</td></tr><tr><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">Missing Tables</td><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal"><span style="color:#905;background-color:#ddd">`r missing_tables`</span></td></tr><tr><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">Total number of unique nominators</td><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal"><span style="color:#905;background-color:#ddd">`r nrow(overall_nominators)`</span></td></tr></tbody></table>

# Changes over time
This section investigates the changes over time in the bonded amount and the targets of each individual nominator.

### Bonded Amount: Histogram of changes

The analysis focuses on the frequency of changes in the bonded amount. A change is only effective after a new election, which is held every `r number_session_per_era` sessions. Therefore, it makes sense to normalize the frequency to "per era" (instead of per session). In other words, if a nominator changes their bonded amount once over `r number_session_per_era` sessions, the frequency in the histogram would be 1. If a nominator changes their bonded_amount each session, it would be counted as 6.

```{r, echo = FALSE}
# Create histogram with ggplot2
ggplot(overall_nominators, aes(x = bonded_changed_percent)) +
  geom_histogram(binwidth = 1, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Bonded Amount Changed Percentage",
       x = "Bonded Amount Changed Percentage", y = "Frequency")
```

There are some outliers of nominators that very frequently adjust their bonded amount. Let's focus on those that lie between 0 and 1.

```{r, echo = FALSE}
# Create histogram with ggplot2
ggplot(subset(overall_nominators, bonded_changed_percent <= 1), aes(x = bonded_changed_percent)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Bonded Amount Changed Percentage",
       x = "Bonded Amount Changed Percentage", y = "Frequency")
```

### Bonded amount: Changes over sessions

The following graph plots the changes in bonded amount over the sessions. The frequency is in absolute numbers, i.e., how many nominators changed their bonded_amount within each session. 

```{r, echo = FALSE}
ggplot(subset(bonded_changed_per_session_aggregated, session != first_session), aes(x = session, y = average_nr_changes)) +
  geom_line() +
  geom_point(size = 0.5) +
  theme_minimal() +
  labs(title = "Number of Changes Over Time (Bonded Amount Change)",
       x = "Session",
       y = "Number of Changes") +
  theme(panel.grid.minor = element_blank())
```

We can observe a high number of changes in the bonded amount, but this is mainly due to the fact that many nominators utilize the automatic re-bonding of staking rewards. It is difficult to distinguish manual changes of bonded amount from automatic ones. This could be approximated, however, given the size of the stash address and their resulting expected staking rewards per session. This is left open for now.


## Targets
This analysis is the most important part of nominator's behavior as it gives insights directly about their validator preferences and the frequency of updating their nominations. The following histogram plots the individual changes in targets over the nominators over the whole time frame. It is normalized over the number of sessions a nominator had been existent and per era. In other words, a value of 1 would mean that a nominator changed their targets once every six sessions that they were existent. Note that, compared to the `bonded_amount`, a change in targets is always manually triggered by a nominator and therefore is a a conscious behavior.

### Targets: Histogram of changes

```{r, echo = FALSE}
# Create histogram with ggplot2
ggplot(overall_nominators, aes(x = targets_changed_percent)) +
  geom_histogram(binwidth = 1, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Targets Changed Percentage",
       x = "Targets Changed Percentage", y = "Frequency")
```

Again, we have some outliers of nominators that very frequently change their targets, which decreases the readability of the graph. The following graph focuses on nominators that change their targets every six sessions between 0 and 100% of times.

```{r, echo = FALSE}
# Create histogram with ggplot2
ggplot(subset(overall_nominators, targets_changed_percent <= 1), aes(x = targets_changed_percent)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Targets Changed Percentage",
       x = "Targets Changed Percentage", y = "Frequency")
```

The following graph plots the frequency of target changes of nominators per session.

```{r, echo = FALSE}
ggplot(subset(targets_changed_per_session_aggregated, session != first_session), aes(x = session, y = average_nr_changes)) +
  geom_line() +
  geom_point(size = 0.5) +
  theme_minimal() +
  labs(title = "Number of Changes Over Time (Targets Change)",
       x = "Session",
       y = "Number of Changes") +
  theme(panel.grid.minor = element_blank())
```



# Nomination of inactive validators

```{r, echo = FALSE}
newest_nominators <- polkadotutils::getStakingData("nominators", "session", last_session, chain, "online")
newest_validators <- polkadotutils::getStakingData("validators", "session", last_session, chain, "online")

# Unnest the targets column for each nominator and create a new data frame that has a row for each entry in "targets".
nominators_unnested <- subset(newest_nominators, select = c("stash_address", "targets")) %>%
    separate_rows(targets, sep = ",")

colnames(nominators_unnested) <- c("nominator_stash", "stash_address")

merged <- merge(nominators_unnested, subset(newest_validators, select = c("stash_address", "active")), by = "stash_address", all = TRUE)
merged$counter <- 1

merged_aggregated <- aggregate(cbind(active, counter) ~ nominator_stash, data = merged, FUN = sum, na.rm = TRUE)

merged_aggregated$ratio <- merged_aggregated$active / merged_aggregated$counter
merged_aggregated$at_least_one_inactive <- ifelse(merged_aggregated$ratio < 1, 1,0)
  
```

The following histogram plots the ratio of active to inactive validators in the targets of the nominators. In contrast to the previous analyses, this investigates the current state of the network of the newest session (`r last_session`). Here, a ratio of 1 means that all the validators that a nominator is voting for are currently active.

```{r, echo=FALSE}
ggplot(merged_aggregated, aes(x = ratio)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "lightblue") +
  labs(title = "Histogram of ratio of nominations of active validators",
       x = "Ratio of nominations of active validators", y = "Frequency")
```

In general, there are `r format(sum(merged_aggregated$at_least_one_inactive), scientific = FALSE)` (or `r round(sum(merged_aggregated$at_least_one_inactive) / nrow(merged_aggregated)*100, digits=2)`%) nominators that have at least one inactive validator in their targets.
