---
title: "Nominators' Behavior"
author: "Jonas Gehrlein @ Web3 Foundation"
date: 20/04/2023
output: html_document
version: "1.0"
---


```{r setup, fig.width=4, fig.height=3, warning=FALSE, message=FALSE, echo=FALSE}
# Load the required R packages:
library(plyr)
library(dplyr)
library(rmarkdown)
library(ggplot2)
library(reshape2)
library(tidyverse)
library(scales)
library(dbplyr)
library(assert)
library(ineq)
library(ggraph)
library(splitstackshape)
library(RCurl)
library(polkadotutils)
library(data.table)
library(googleCloudStorageR)
library(bigrquery)
library(httr)
library(data.table)
library(bit64)
```


```{r}
# Set the OAuth scopes
options(bigrquery.oauth.scopes = c("https://www.googleapis.com/auth/cloud-platform"))
bigrquery::bq_auth(email = "jonas@web3.foundation", cache = TRUE)

# Authenticate googleCloudStorageR using the token obtained by bigrquery
gcs_auth(token = bq_token())
```

```{r, message=FALSE, echo = FALSE}
# User Input
# Can be "polkadot" or "kusama"
chain = "polkadot"

# if we want not every era but every `time_offset` era. Using "1" means we include all eras. Using "2" would mean every second era etc.
time_offset <- 7*6*36
# The amount of tables that should be downloaded
number_tables <- 5
# The sessions to analyze.
last_session = 9325
first_session = last_session - time_offset * number_tables

missing_tables <- 0

```

```{r, echo=FALSE}
# Parameters to determine whether a change in bonded amount was due to auto-compound or an active choice. Here, we pretend that an increase by bonded_amount * APY / 365 is a compound. I.e., if the change from old to new balance is less than that, we pretend its not an active choice. Making APY higher increases the range where we pretend it was an auto compound.
APY <- 0.18
number_session_per_era <- 6  # Payout happens every 6 sessions
payout_periods <- ceiling(time_offset / number_session_per_era)  # How many payout periods are covered by time_offset
threshold <- 1 + payout_periods * (APY / 365)
```

```{r, echo=FALSE, message=FALSE}
#Setting up the denomination KSM and (new) DOT:
network_identifier = ifelse(chain == "polkadot", "DOT", "KSM")

if(chain=="polkadot"){
  normalization = 1/10000000000
  chain_name = "Polkadot"
} else {
  normalization = 1/1000000000000
  chain_name = "Kusama"
}

# Additional normalization
normalization_thousands = 1/1000
normalization_millions = 1/1000000
normalization_billions = 1/1000000000

difference = last_session - first_session 
#x = c(last_session:(last_session - difference))

sessions = c(first_session:(first_session + difference))
# extract every time_offset's value
sessions <- sessions[seq(1, length(sessions), time_offset)]
old <- Sys.time() # get start time
```

```{r, echo=FALSE}
initiate_data <- function(df_t0, df_t1, threshold) {
  # Full outer join between df_t0 and df_t1 by stash_address
  merged_df <- merge(df_t0, df_t1, by = "stash_address", all = TRUE, suffixes = c("_t0", "_t1"))

  # Initialize the session_count based on whether session_t0 and session_t1 exist
  merged_df$session_count <- ifelse(is.na(merged_df$session_t0) & !is.na(merged_df$session_t1), 1,
                               ifelse(!is.na(merged_df$stash_address) & is.na(merged_df$session_t1), 1, 2))
  
  # Vectorized logic to identify targets_changed
  merged_df$targets_changed <- mapply(function(t0, t1, s0, s1, df_t1_session) {
    if (is.na(s0) && !is.na(s1)) {
      return(list(s1))
    } else if (!is.na(s0) && is.na(s1)) {
      return(list(c(s0, df_t1_session)))
    } else if (!is.na(t0) && t0 != t1) {
      return(list(c(s0, s1)))
    } else {
      return(list(s0))
    }
  }, merged_df$targets_t0, merged_df$targets_t1, merged_df$session_t0, merged_df$session_t1, MoreArgs = list(df_t1$session[1]))

  # Vectorized logic to identify bonded_amount_changed
  merged_df$bonded_amount_changed <- mapply(function(b0, b1, s0, s1, df_t1_session) {
    if (is.na(s0) && !is.na(s1)) {
      return(list(s1))  # New entry, track the new session
    } else if (!is.na(s0) && is.na(s1)) {
      return(list(c(s0, df_t1_session)))  # Old entry dropped, track the session
    } else if (!is.na(b0) && b1 > b0) {
      if (b1 > b0 * threshold) {
        return(list(c(s0, s1)))  # Trigger change if the increase exceeds the threshold
      } else {
        return(list(s0))  # No significant change
      }
    } else if (!is.na(b0) && b1 < b0) {
      return(list(c(s0, s1)))  # Balance decreased, track the session
    } else {
      return(list(s0))  # No change
    }
  }, 
  merged_df$bonded_amount_t0, 
  merged_df$bonded_amount_t1, 
  merged_df$session_t0, 
  merged_df$session_t1, 
  MoreArgs = list(df_t1$session[1])
)

  # Create the final data frame
  final_df <- data.frame(
    stash_address = merged_df$stash_address,
    targets = ifelse(is.na(merged_df$session_t1), NA, 
                     ifelse(is.na(merged_df$targets_t1), merged_df$targets_t0, merged_df$targets_t1)),
    bonded_amount = ifelse(is.na(merged_df$session_t1), NA, 
                           ifelse(is.na(merged_df$bonded_amount_t1), merged_df$bonded_amount_t0, merged_df$bonded_amount_t1)),
    targets_changed = I(merged_df$targets_changed),
    bonded_amount_changed = I(merged_df$bonded_amount_changed),
    session_count = merged_df$session_count,
    session = merged_df$session_t1
  )

  return(final_df)
}
```

```{r, echo = FALSE}
update_data <- function(initial_data, new_data, threshold) {
  # Ensure that stash_address is unique in both data sets
  initial_data <- initial_data[!duplicated(initial_data$stash_address), ]
  new_data <- new_data[!duplicated(new_data$stash_address), ]

  # Full outer join (all = TRUE) to retain all nominators
  merged_df <- merge(initial_data, new_data, by = "stash_address", all = TRUE, suffixes = c("_old", "_new"))

  # Vectorized update of targets_changed
  merged_df$targets_changed <- mapply(function(t_old, t_new, s_old, s_new, changed_list, new_session) {
    if (!is.na(s_old) && !is.na(s_new)) {
      if (t_old != t_new) {
        if (!(s_new %in% changed_list)) {
          return(append(changed_list, s_new))  # Add session if target changed
        } else {
          return(changed_list)
        }
      } else {
        return(changed_list)
      }
    } else if (is.na(s_old) && !is.na(s_new)) {
      return(c(s_new))  # New session entry
    } else if (!is.na(s_old) && is.na(s_new)) {
      return(append(changed_list, new_session))  # Old session entry dropped
    } else {
      return(changed_list)
    }
  }, merged_df$targets_old, merged_df$targets_new, merged_df$session_old, merged_df$session_new, merged_df$targets_changed, MoreArgs = list(new_data$session[1]))

  # Vectorized update of bonded_amount_changed
  merged_df$bonded_amount_changed <- mapply(function(b_old, b_new, s_old, s_new, changed_list, new_session) {
    if (!is.na(s_old) && !is.na(s_new)) {
      if (b_new > b_old * threshold) {
        if (!(s_new %in% changed_list)) {
          return(append(changed_list, s_new))  # Trigger change if balance increased and exceeded threshold
        } else {
          return(changed_list)
        }
      } else if (b_new < b_old) {
        if (!(s_new %in% changed_list)) {
          return(append(changed_list, s_new))  # Trigger change if balance decreased
        } else {
          return(changed_list)
        }
      } else {
        return(changed_list)  # No significant change
      }
    } else if (is.na(s_old) && !is.na(s_new)) {
      return(c(s_new))  # New session entry
    } else if (!is.na(s_old) && is.na(s_new)) {
      return(append(changed_list, new_session))  # Old session entry dropped
    } else {
      return(changed_list)
    }
  }, merged_df$bonded_amount_old, merged_df$bonded_amount_new, merged_df$session_old, merged_df$session_new, merged_df$bonded_amount_changed, MoreArgs = list(new_data$session[1]))

  # Update the session count
  merged_df$session_count <- ifelse(!is.na(merged_df$session_old) & !is.na(merged_df$session_new),
                                    merged_df$session_count + 1,
                                    ifelse(is.na(merged_df$session_old) & !is.na(merged_df$session_new), 1, merged_df$session_count))

  # Create the final data frame
  final_df <- data.frame(
    stash_address = merged_df$stash_address,
    targets = ifelse(is.na(merged_df$session_new), merged_df$targets_old, merged_df$targets_new),
    bonded_amount = ifelse(is.na(merged_df$session_new), merged_df$bonded_amount_old, merged_df$bonded_amount_new),
    targets_changed = I(merged_df$targets_changed),
    bonded_amount_changed = I(merged_df$bonded_amount_changed),
    session_count = merged_df$session_count,
    session = merged_df$session_new,
    stringsAsFactors = FALSE
  )

  return(final_df)
}
``` 

```{r}
calculate_days_since_change <- function(overall_nominators, session_timestamps) {
  # Get the timestamp of the most recent session as the reference
  latest_session <- max(as.numeric(names(session_timestamps)))
  reference_timestamp <- session_timestamps[[as.character(latest_session)]]

  # Calculate days since last bonded amount change
  overall_nominators$days_since_bonded_amount_changed <- sapply(overall_nominators$bonded_amount_changed, function(changed_list) {
    # Find the highest session in the bonded_amount_changed list
    if (length(changed_list) > 0) {
      last_change_session <- max(unlist(changed_list))
      # Look up the timestamp of the highest session
      last_change_timestamp <- session_timestamps[[as.character(last_change_session)]]
      if (!is.null(last_change_timestamp)) {
        # Calculate the difference in days between the last change timestamp and the reference timestamp
        as.numeric(difftime(as.POSIXct(reference_timestamp, origin = "1970-01-01", tz = "UTC"), 
                            as.POSIXct(last_change_timestamp, origin = "1970-01-01", tz = "UTC"), 
                            units = "days"))
      } else {
        return(NA)
      }
    } else {
      return(NA)
    }
  })

  # Calculate days since last target change
  overall_nominators$days_since_target_changed <- sapply(overall_nominators$targets_changed, function(changed_list) {
    # Find the highest session in the targets_changed list
    if (length(changed_list) > 0) {
      last_change_session <- max(unlist(changed_list))
      # Look up the timestamp of the highest session
      last_change_timestamp <- session_timestamps[[as.character(last_change_session)]]
      if (!is.null(last_change_timestamp)) {
        # Calculate the difference in days between the last change timestamp and the reference timestamp
        as.numeric(difftime(as.POSIXct(reference_timestamp, origin = "1970-01-01", tz = "UTC"), 
                            as.POSIXct(last_change_timestamp, origin = "1970-01-01", tz = "UTC"), 
                            units = "days"))
      } else {
        return(NA)
      }
    } else {
      return(NA)
    }
  })

  return(overall_nominators)
}
```


```{r, echo=FALSE}
# To get a vector of sessions that were found (excluding those that have no data)
# Getting the first data frame to start the process
nominators_t0 <- getStakingDataNew("nominators", "session", sessions[1], chain, "online", TRUE)

session_timestamps <- list()

# Manually add the first session timestamp from nominators_t0
session_timestamps[[as.character(nominators_t0$session[1])]] <- nominators_t0$timestamp[1] / 1000  # Convert from milliseconds to seconds
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
for (i in 2:length(sessions)) {
  message(paste0("Processing session: ", sessions[i]))
  
  # Try to download the next session's nominator data
  nominators_next <- tryCatch(
    {
      getStakingDataNew("nominators", "session", sessions[i], chain, "online", TRUE)
    },
    error = function(e) {
      message(paste("Error in session", sessions[i], "- skipping. Error message:", e$message))
      missing_tables <<- missing_tables + 1  # Increment the missing tables counter
      return(NULL)  # Return NULL to indicate failure
    }
  )
  
  # Skip this iteration if nominators_next is NULL (data couldn't be downloaded)
  if (is.null(nominators_next)) {
    next
  }
  session_timestamps[[as.character(nominators_next$session[1])]] <- nominators_next$timestamp[1] / 1000  # Convert from milliseconds to seconds

  # If we are in the second iteration, initialize the data
  if (i == 2) {
    overall_nominators <- initiate_data(nominators_t0, nominators_next, threshold)
  } else {
    overall_nominators <- update_data(overall_nominators, nominators_next, threshold)
  }
  
  # Save intermediate results to avoid excessive memory usage
  if (i %% 10 == 0) {
    saveRDS(overall_nominators, file = paste0("overall_nominators_session_", i, ".rds"))
    message(paste("Saved intermediate results at session", i))
  }
  
  # Explicitly call garbage collection to free memory
  gc()

  # Optionally, remove intermediate data
  rm(nominators_next)
}

# Final message indicating how many tables were missing
message(paste("Total missing tables:", missing_tables))
# No need to keep the expensive column
overall_nominators <- overall_nominators %>%
  select(-targets)

# print elapsed time
new <- Sys.time() - old # calculate difference
print(new) # print in nice format
```

```{r}
overall_nominators <- calculate_days_since_change(overall_nominators, session_timestamps)

# Add an indicator to the full list of all nominators ever present of which are the current nominators.
overall_nominators <- overall_nominators %>%
  mutate(current_set = if_else(stash_address %in% newest_nominators$stash_address, 1, 0))
```

```{r, echo=FALSE}
# ------------------------ General data manipulation --------------------------

## Bonded Amount changes
# Getting the number of entries in the list to know how many changes occurred
for(i in 1:nrow(overall_nominators)){
  overall_nominators$nr_amount_changed[i] <- length(overall_nominators$bonded_amount_changed[i][[1]])
}

# Getting the number of entries in the list to know how many changes occurred
for(i in 1:nrow(overall_nominators)){
  overall_nominators$nr_targets_changed[i] <- length(overall_nominators$targets_changed[i][[1]])
}

overall_nominators$total_changes <- overall_nominators$nr_amount_changed + overall_nominators$nr_targets_changed
```

```{r echo = FALSE}
## Analysis Targets Change

# Create a dataframe that unnests the list of sessions per nominator and aggregates the number of changes on session level
targets_changed_per_session <- overall_nominators %>%
  select(stash_address, targets_changed) %>%
  unnest(cols = targets_changed)
targets_changed_per_session$tracker <- 1
targets_changed_per_session_aggregated <- aggregate(targets_changed_per_session$tracker, by = list(targets_changed_per_session$targets_changed), FUN = sum)
colnames(targets_changed_per_session_aggregated) <- c("session", "number_changes")

# There are a few missing tables which means that changes "aggregate" over a longer period of time until a new data set is found and then all the changes are accounted to that new session. To smoothen this out, I take the average change per session if there are missing tables. 
for(i in 1:nrow(targets_changed_per_session_aggregated)){
  if(i == 1){
    targets_changed_per_session_aggregated$distance[i] <- 1
  } else {
    targets_changed_per_session_aggregated$distance[i] <- targets_changed_per_session_aggregated$session[i] - targets_changed_per_session_aggregated$session[i-1]
  }
}
targets_changed_per_session_aggregated$average_nr_changes <- targets_changed_per_session_aggregated$number_changes / targets_changed_per_session_aggregated$distance
# The percentage of changes in the targets of a nominator. Since we count the first appearance of a nominator as a change in target, we substract it here. And since this is session data, a reasonable time frame to change something would be once per era at maximum. 
overall_nominators$targets_changed_percent <- (overall_nominators$nr_targets_changed - 1) / (overall_nominators$session_count/number_session_per_era)
```

```{r, echo = FALSE}
# Analysis Bonded Amount Change
bonded_changed_per_session <- overall_nominators %>%
  select(stash_address, bonded_amount_changed) %>%
  unnest(cols = bonded_amount_changed)

bonded_changed_per_session$tracker <- 1
bonded_changed_per_session_aggregated <- aggregate(bonded_changed_per_session$tracker, by = list(bonded_changed_per_session$bonded_amount_changed), FUN = sum)
colnames(bonded_changed_per_session_aggregated) <- c("session", "number_changes")

for(i in 1:nrow(bonded_changed_per_session_aggregated)){
  if(i == 1){
    bonded_changed_per_session_aggregated$distance[i] <- 1
  } else {
    bonded_changed_per_session_aggregated$distance[i] <- bonded_changed_per_session_aggregated$session[i] - bonded_changed_per_session_aggregated$session[i-1]
  }
}
bonded_changed_per_session_aggregated$average_nr_changes <- bonded_changed_per_session_aggregated$number_changes / bonded_changed_per_session_aggregated$distance
# The percentage of changes in the bonds of a nominator. Since we count the first appearance of a nominator as a change in bonds, we subtract it here. And since this is session data, a reasonable time frame to change something would be once per era at maximum. 
overall_nominators$bonded_changed_percent <- (overall_nominators$nr_amount_changed - 1) / (overall_nominators$session_count/number_session_per_era)
```

# Introduction

This analysis gives insights into the behavior of nominators within the `r chain` network. Nominators are tracked across consecutive sessions using their unique stash address. Within this dataset, each session provides insight into the `bonded_amount` and a list of targets, representing the nominators' selected nominations. This approach enables a comparison between sessions, effectively monitoring any changes that occur over time.

The following table illustrates the parameters of the analysis. Note, that there have been some outages in the data collection, but the analysis copes with that and minimizes the impact on the outcome quality.
<table style="border-collapse:collapse;border-spacing:0" class="tg"><thead><tr><th style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;font-weight:normal;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">Parameter</th><th style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;font-weight:normal;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">Value</th></tr></thead><tbody><tr><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">Chain</td><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">`r chain`</td></tr><tr><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">First Session</td><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">`r first_session`</td></tr><tr><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">Last Session</td><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">`r last_session`</td></tr><tr><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">Missing Tables</td><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal"><span style="color:#905;background-color:#ddd">`r missing_tables`</span></td></tr><tr><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal">Total number of unique nominators</td><td style="border-color:inherit;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;text-align:left;vertical-align:top;word-break:normal"><span style="color:#905;background-color:#ddd">`r nrow(overall_nominators)`</span></td></tr></tbody></table>

# Changes over time
In this section, we examine the temporal fluctuations in both the bonded amount and the targets chosen by each nominator. This analysis helps to identify trends and patterns in the behavior of nominators over time.

### Bonded Amount: Histogram of changes

The analysis primarily emphasizes the frequency of alterations in the bonded amount. It is essential to note that changes only take effect following a new election, which occurs every `r number_session_per_era` sessions. To account for this, it is more meaningful to normalize the frequency on a "per era" basis rather than per session. In this context, if a nominator modifies their bonded amount once throughout `r number_session_per_era` sessions, the frequency in the histogram is represented as 1. Conversely, if a nominator adjusts their bonded_amount every session, the frequency is denoted as 6.

```{r, echo = FALSE}
# Create histogram with ggplot2
ggplot(overall_nominators, aes(x = bonded_changed_percent)) +
  geom_histogram(binwidth = 1, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Bonded Amount Changed Percentage",
       x = "Bonded Amount Changed Percentage", y = "Frequency")
```

There are a few outlier nominators who adjust their bonded amount with remarkable frequency. For the purpose of this analysis, we will concentrate on those with a frequency lying between 0 and 1, ensuring a more representative understanding of typical nominator behavior.

```{r, echo = FALSE}
# Create histogram with ggplot2
ggplot(subset(overall_nominators, bonded_changed_percent <= 1), aes(x = bonded_changed_percent)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Bonded Amount Changed Percentage",
       x = "Bonded Amount Changed Percentage", y = "Frequency")
```

### Bonded amount: Changes over sessions

The subsequent graph illustrates the variations in bonded amount across sessions. The frequency displayed is in absolute terms, representing the total number of nominators who altered their bonded_amount within each individual session. This visualization helps to convey the overall trends and patterns in nominators' behavior over time.

```{r, echo = FALSE}
ggplot(subset(bonded_changed_per_session_aggregated, session != first_session), aes(x = session, y = average_nr_changes)) +
  geom_line() +
  geom_point(size = 0.5) +
  theme_minimal() +
  labs(title = "Number of Changes Over Time (Bonded Amount Change)",
       x = "Session",
       y = "Number of Changes") +
  theme(panel.grid.minor = element_blank())
```

We can observe a high number of changes in the bonded amount. However, this is primarily due to many nominators employing automatic re-bonding of staking rewards. Distinguishing manual changes in the bonded amount from automatic ones is challenging. An approximation could be made based on the stash address size and the expected staking rewards per session, but this is left unexplored for now.

## Targets
This analysis forms the most crucial aspect of nominator behavior as it directly sheds light on their validator preferences and the frequency with which they update their nominations. The following histogram displays individual changes in targets across nominators over the entire time frame. The data is normalized by the number of sessions a nominator has been active and is presented on a per-era basis. In other words, a value of 1 signifies that a nominator altered their targets once every six sessions they were active. It is important to note that, in contrast to the `bonded_amount`, changes in targets are always manually initiated by nominators, reflecting conscious decision-making.

### Targets: Histogram of changes

```{r, echo = FALSE}
# Create histogram with ggplot2
ggplot(overall_nominators, aes(x = targets_changed_percent)) +
  geom_histogram(binwidth = 1, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Targets Changed Percentage",
       x = "Targets Changed Percentage", y = "Frequency")
```

Once more, we encounter some outlier nominators who alter their targets with exceptional frequency, which hampers the graph's readability. To address this, the following graph narrows its focus to nominators who modify their targets between 0 and 100% of the time, specifically every six sessions. This refined visualization aids in better understanding the typical behavior of nominators in relation to their target changes.

```{r, echo = FALSE}
# Create histogram with ggplot2
ggplot(subset(overall_nominators, targets_changed_percent <= 1), aes(x = targets_changed_percent)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Targets Changed Percentage",
       x = "Targets Changed Percentage", y = "Frequency")
```

The following graph plots the frequency of target changes of nominators per session.

```{r, echo = FALSE}
ggplot(subset(targets_changed_per_session_aggregated, session != first_session), aes(x = session, y = average_nr_changes)) +
  geom_line() +
  geom_point(size = 0.5) +
  theme_minimal() +
  labs(title = "Number of Changes Over Time (Targets Change)",
       x = "Session",
       y = "Number of Changes") +
  theme(panel.grid.minor = element_blank())
```



# Nomination of inactive validators

```{r, echo = FALSE}
newest_nominators <- polkadotutils::getStakingDataNew("nominators", "session", last_session, chain, "online", TRUE)
newest_validators <- polkadotutils::getStakingDataNew("validators", "session", last_session, chain, "online", TRUE)

# Unnest the targets column for each nominator and create a new data frame that has a row for each entry in "targets".
nominators_unnested <- subset(newest_nominators, select = c("stash_address", "targets")) %>%
    separate_rows(targets, sep = ",")

colnames(nominators_unnested) <- c("nominator_stash", "stash_address")

merged <- merge(nominators_unnested, subset(newest_validators, select = c("stash_address", "active")), by = "stash_address", all = TRUE)
merged$counter <- 1

merged_aggregated <- aggregate(cbind(active, counter) ~ nominator_stash, data = merged, FUN = sum, na.rm = TRUE)

merged_aggregated$ratio <- merged_aggregated$active / merged_aggregated$counter
merged_aggregated$at_least_one_inactive <- ifelse(merged_aggregated$ratio < 1, 1,0)
  
```

The subsequent histogram presents the ratio of active to inactive validators within the nominators' targets. Unlike previous analyses, this examination focuses on the current state of the network during the most recent session (`r last_session`). In this context, a ratio of 1 indicates that all validators a nominator is voting for are presently active. This insight offers an understanding of the overall alignment of nominator preferences with the active validator pool.

```{r, echo=FALSE}
ggplot(merged_aggregated, aes(x = ratio)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "lightblue") +
  labs(title = "Histogram of ratio of nominations of active validators",
       x = "Ratio of nominations of active validators", y = "Frequency")
```

In general, there are `r format(sum(merged_aggregated$at_least_one_inactive), scientific = FALSE)` (or `r round(sum(merged_aggregated$at_least_one_inactive) / nrow(merged_aggregated)*100, digits=2)`%) nominators that have at least one inactive validator in their targets.
